Determining the existance and direction of causality in the face of measurement error

### 2016-05-05 docs/manuscript.tex
updated

\setlength{\droptitle}{-2em}
  \title{Determining the existance and direction of causality in the face of
measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{25 February 2016}

@@

* Correspondence to:
\href{mailto:g.hemani@bristol.ac.uk}{\nolinkurl{g.hemani@bristol.ac.uk}}

@@

With the explosion in data that assess the human phenome, we are
presented with the challenge of understanding the causal nature of the
emerging putitive correlations between measures of interest. Mediation
is a commonly used statistical method for making causal inference in
observational data\textsuperscript{1--4}, and it exists in many forms,
from simple regression based systems to structural equation modeling
(SEM)\textsuperscript{5}. The principle behind mediation analysis can be
explained as follows. Supposing an exposure has a causal effect on an
outcome, then an `instrument' that causes the exposure (a single
nucleotide polymorophism (SNP), for example) should also influence the
outcome. Therefore the influence of the instrument on the outcome
conditional on the exposure should be zero. This forms the basis of a
number of methods, such as the regression-based causal inference test
(CIT)\textsuperscript{1}, an SEM implementation in the NEO
software\textsuperscript{5}, and various Bayesian methods, which have
been employed by a number of recent publications that make causal
inferences, often in large scale `omics datasets\textsuperscript{6--8}.

While mediation based approaches are simple and convenient to implement,
it is important to note where they may lead to unreliable results. One
such mechanism is measurement error\textsuperscript{9--12}, which has
been noted to be an explicit issue with CIT style
approaches\textsuperscript{13,14}. Measurement (or observational) error
is the difference between a measured value of a quantity and its true
value. Such variability can arise through a whole plethora of
mechanisms, which are often unique to the study design and difficult to
avoid\textsuperscript{15,16}. Array technology is now commonly used to
obtain high throughput phenotyping at low cost, but they come with the
problem of having imperfect sensitivity, so for example methylation
levels as measured by the Illumina450k chip are prone to have some
amount of noise around the true value\textsuperscript{17,18}.
Sensitivity is also an issue, for example if the unit of measurement of
biological interest is the methylation level in a T cell, then
measurement error of this value can be introduced by using methylation
levels from whole blood samples because the measured value will be an
assay of many cell types\textsuperscript{19}.

Measurement error will of course arise in more low-tech data too, for
example when measuring body mass index (BMI) one is typically interested
in using this as a proxy for obesity, but it is clear that the
correlation between BMI and underlying adiposity is not
perfect\textsuperscript{20}. A similar problem of biological
misspecification is unavoidable in disease diagnosis. Measurement error
can also be introduced after the data has been collected, for example
the transformation of non-normal data for the purpose of statistical
analysis will lead to a new variable that will typically have both bias
and imprecision compared to the original variable. The sources of
measurement error are not limited to this list\textsuperscript{16}, and
its impact has been explored in the context of mediation analysis in the
epidemiological literature extensively\textsuperscript{9,12}.

An alternative statistical method that gained prominence as a solution
to the problem of measurement error is Mendelian randomisation
(MR)\textsuperscript{21,22}. Here an instrumental variable (typically a
SNP) that has a robost, causal association with the exposure is used to
stand for the exposure in an association test with the outcome. If there
is no measurement error in the SNP then the causal association between
the exposure and the outcome can be estimated by scaling the association
between the SNP and the outcome by the association between the SNP and
the exposure. Indeed this approach was in part pioneered (in the form of
instrumental variable (IV) analysis) as a correction for measurement
error\textsuperscript{23}, but it also has the useful property that it
guards against unmeasured confounders driving the association between
exposure and outcome; and if the biological effect of the SNP is
understood then it can also aid against reverse causality driving the
association.

Here, however, arises a potential issue: often the biological effect of
a SNP is not known. If there exists a putative association between two
variables, with a SNP being robustly associated with each, it can be
difficult to determine which of the two variables is subject to the
direct effect of the SNP (i.e., for which of the two variables is the
SNP a valid instrument?). By definition, we expect that if the
association is causal then a SNP for the exposure will be associated
with the outcome, so if the researcher erroneously uses the SNP as an
instrument for the outcome then they are likely to see an apparently
robust causal association of outcome on exposure. Such a situation can
arise in many scenarios. For example genome wide association studies
(GWASs) that identify genetic associations for complex traits are, by
design, hypothesis free and agnostic of genomic function, and it often
takes years of follow up studies to understand the biological nature of
a putative GWAS hit\textsuperscript{24}. To illustrate a trivial example
- the same FTO variant that is famously associated with BMI is also
associated with C-reactive protein (CRP). It is known that this is
because BMI causes CRP\textsuperscript{25}, but supposing we only had
\emph{a priori} knowledge that FTO associated with CRP, and we used it
to instrument CRP in an MR analysis of CRP against BMI, we would infer
that CRP causes BMI. Another situation is where the causal direction
between 'omic measures need to be determined, for example if a DNA
methylation probe is associated with expression of an adjacent gene,
then is a cis-acting SNP an instrument for the DNA methylation level, or
the gene expression level\textsuperscript{8}?

@@

The analysis proceeds by simulating variables that have known causal
effects and valid instruments, and then applying measurement error in
different configurations. We then assess the performance of the CIT test
and MR analysis for a) inference of a causal relationship and b)
inferring the correct direction of causality. Included in the MR
analysis is a simple extension that is used to make inference of the
causal direction when it is unknown which of the correlated variables
the instrument is directly acting upon. All analysis was performed using
the R programming language\textsuperscript{26} and code is made
available at \href{}{github url to go here}.

@@

The CIT method\textsuperscript{1} is implemented in the R package
\emph{R/cit}\textsuperscript{27}. The \emph{cit.cp} function was used to
obtain an omnibus p-value of causality between an exposure \(x\) and an
outcome \(y\) using a mediator \(g\) by making inferences based on a
number of conditions that we examine in the results. To infer the
direction of causality using the CIT method, an omnibus p-value
generated by CIT, \(p_{CIT}\), was estimated for the correct model of
\(x\) causing \(y\), and for incorrect model of \(y\) causing \(x\). For
some significance threshold \(\alpha\), causal direction was inferred
based on the following scenarios:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  If \(p_{CIT,correct} < \alpha\) and \(p_{CIT,incorrect} > \alpha\)
  then the correct model is accepted
\item
  If \(p_{CIT,correct} > \alpha\) and \(p_{CIT,incorrect} < \alpha\)
  then the incorrect model is accepted
\item
  If \(p_{CIT,correct} > \alpha\) and \(p_{CIT,incorrect} > \alpha\)
  then the an alternative causal model is accepted
\item
  If \(p_{CIT,correct} < \alpha\) and \(p_{CIT,incorrect} < \alpha\)
  then no inference is made
\end{enumerate}

For the purposes of compiling simulation results we use an arbitrary
\(\alpha = 0.05\) value.

@@

Supposing there are two correlated variables, \(x\) and \(y\), with a
variant \(g\) associated with both. Assuming no horizontal pleiotropy
(i.e. \(g\) only has an effect on one of the variables through the other
variable) it is desirable to know which of the variables it has a direct
influence on. This can be achieved by testing for a difference in the
correlations \(\rho_{g, x}\) and \(\rho_{g, y}\) using Steiger's Z-test
for correlated correlations within a population\textsuperscript{28}. It
is calculated as

@@

where Fisher's z-transformation is used to obtain
\(Z_{g*} = \frac{1}{2} \ln \left ( \frac{1+\rho_{g*}}{1-\rho_{g*}} \right )\),

@@

The \(Z\) value is interpreted such that

@@

and a p-value is generated from the \(Z\) value to indicate confidence
of difference in correlations \(\rho_{gx}\) and \(\rho_{gy}\). The test
for a causal association is then obtained by performing two sample MR in
the direction that is predicted to be the correct model based on the Z
score test. For the purposes of compiling simulation results, the test
is deemed to have obtained a robust association if the Z score test has
a p-value \(< 0.05\) and the two sample MR analysis has a p-value
\(< 0.05\), though in practice we do not advocate using an arbitrary
\(p\) value threshold for denoting significance in making causal
inference.

Note that the same correlation test approach can be applied to a
two-sample MR\textsuperscript{29} setting, where the Steiger test of two
independent correlations is applied instead of the case detailed here
where correlations share one variable in common. An advantage of using
the Steiger test in the two sample context is that it can compare
correlations in independent samples where sample sizes are different.

@@

Simulations were conducting by creating variables of sample size \(n\)
for the exposure \(x\), the measured values of the exposure \(x_O\), the
outcome \(y\), the measured values of the outcome \(y_O\) and the
instrument \(g\). In all models \(x\) causes \(y\) and \(g\) is an
instrument for \(x\). Each variable was simulated such that:

@@

All \(\beta\) values were set to 1. Values of \(\epsilon_*\) were
generated such that

@@

giving a total of 432 combinations of parameters. Each of these sets of
variables was performed 100 times, and the CIT and MR methods were
applied to each in order to evaluate the causal association of the
simualted variables. Similar patterns of results were obtained for
different values of \(cor(g, x)\).

@@

Measurement error of an exposure can be modeled as some transformation
of the true value that leads to the observed value, \(x_O=f(x)\). For
example, we can define \(f(x) = \alpha_m + \beta_m x + \epsilon_m\),
where \(\alpha_m\) and \(\beta_m\) influence the bias in the measurement
of \(x\), and \(\epsilon_m\) represents the imprecision in the
measurement of \(x\). Here the true value of the exposure is partially
explained by the genetic instrument, \(g\), such that

@@

where \(\beta_g\) is the effect of the SNP on the exposure, and
\(\epsilon_g\) is the residual value of \(x\); and the outcome is
partially explained by the exposure

@@

where \(\beta_x\) is the true effect of the exposure on the outcome. In
the causal inference test (CIT), an omnibus p-value is generated from
four hypothesis tests: 1) \(g\) is associated with \(x\); 2) \(g\) is
associated with \(y\); 3) \(g\) is associated with \(x|y\); and 4) \(g\)
is independent of \(y|x\). The 4th condition is necessary for causal
inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\), where
\(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O\). When measurement error
is introduced we can show using basic covariance properties (Appendix 1)
that

@@

Thus an observational study will find \(cov(g, y - \hat{y}) = 0\) when
the true model is causal only when \(D = 1\). Therefore, if there is any
measurement error that incurs imprecision (i.e.
\(var(\epsilon_m) \neq 0\)) then there will remain an association
between \(g\) and \(y|x\), which is in violation of the the 4th
condition of the CIT. Note that measurement bias alone is insufficient
to lead to a violation of the test statistic assumptions.

We performed simulations to verify that this problem does arise using
the CIT method. Figure 1 shows that when there is no measurement error
in the exposure or outcome variables (\(\rho_{x, x_O}=1\)) the CIT is
reliable in identifying the correct causal direction. However, as
measurement error increases in the exposure variable, eventually the CIT
is more likely to infer a robust causal association in the wrong
direction. Also of concern here is that increasing sample size does not
solve the issue, indeed it only strengthens the evidence of the
incorrect inference.

@@

When selecting an instrument \(g\) that has a direct influence on \(x\)
and \(x\) is causally related to \(y\), one can reason that the
influence of \(g\) on \(y\) is the proportion of the effect of \(x\) on
\(y\) that is explained by the effect of \(g\) on \(x\). Hence the
effect estimate \(\beta_{MR}\) of \(x\) on \(y\) is estimated as

@@

thus we obtain an estimate of the effect of \(x_O\) on \(y\) that is
unbiased by measurement error in \(x\).

However, if we do not know whether the SNP \(g\) has a direct influence
on \(x\) or \(y\) then further efforts are required. If \(x\) causes
\(y\) and \(g\) is a valid instrument for \(x\) then we expect that the
Steiger test will obtain the correct result because the value
\(d = \rho_{g, x_O} - \rho_{g, y_O}\) is greater than 0 when the causal
association between \(x\) and \(y\) is \(\rho_{x, y} < 1\). But it can
be shown that in the presence of measurement error,
\(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\) (Appendix 2), thus it is
important to note that under certain conditions when measurement error
is present the Steiger test could make erroneous inference about causal
direction because the value of \(d\) is not restricted to being greater
than 0. However, using the MR approach there are two potential
advantages over the CIT. First, we can conduct a formal test for the
direction of causality. Second the correct direction of causality can be
inferred when measurement error in the exposure variable is present, as
long as it is lower than the product of the measurement error in the
outcome and the causal correlation between the exposure and the outcome.
Figure 2 shows that in most cases, especially when the causal effect
between \(x\) and \(y\) is not very large, the condition of \(d > 0\) is
satisfied.

We performed simulations to explore the performance of this approach in
comparison to CIT. Figure 3 shows that, as predicted, when \(d < 0\) the
MR analysis is liable to infer the wrong direction of causality, and
that this erroneous result is more likely to occur with increasing
sample size. However, in a large number of cases the MR analysis infers
the wrong direction of causality at a much lower rate than the CIT. When
\(d > 0\) is satisfied we observe that in most cases the MR method has
greater power to obtain evidence for causality than CIT, and always
obtains the correct direction of causality.

@@

Under many circumstances a practical solution to this problem is to use
Mendelian randomisation instead of mediation based methods. Mendelian
randomisation is robust in the face of measurement error and, given that
the researcher has knowledge about the biology of the instrument being
used in the analysis, can offer a direct solution to the issues that CIT
faces. This assumption is often reasonable, for example SNPs are
commonly used as instruments when they are found in genes with known
biological relevance to the trait of interest. But on many occasions
this is not the case, and it may be tempting to resort to using methods
based on mediation in order to be able to both ascertain if there is a
causal association and to infer the direction of causality. Here we have
described a simple extension to MR which can be used as an alternative
to mediation based methods. We show that this method is still liable to
measurement error, but because it has different properties to the CIT it
offers two main advantages. First, it uses a formal statistical
framework to test for the robostness of the assumed direction of
causality. Second, after testing in a comprehensive range of scenarios
the MR based approach is less likely to infer the wrong direction of
causality compared to CIT, while substantially improving power over CIT
in the cases where \(d > 0\).

Mediation based network approaches are very well
established\textsuperscript{14} and have a number of extensions that
make them valuable tools, including for example network construction.
But because they are predicated on the basic underlying principles of
mediation they are liable to suffer from the same issues of measurement
error. Recent advances in MR methodology, for example network
MR\textsuperscript{30} and mediation through MR\textsuperscript{31} may
offer more robust solutions alternatives for these more complicated
problems.

There are a number of limitations to the analysis that we performed. We
focused on the simple case of a single instrument in a single sample
setting, and assumed that pleiotropy (the influence of the instrument on
the outcome through a mechanism other than the exposure) was not
present. To give an example in which this might arise consider the case
that C-reactive protein (CRP) is being tested for causal association
with some outcome, and the FTO variant is used as an instrument. Though
FTO is robustly associated with CRP, it is through the upstream
influence of BMI\textsuperscript{25}. Thus, if BMI has a direct
influence on the outcome, the conditions of MR will be violated and the
Steiger test proposed here will be unreliable.

@@

\includegraphics{manuscript_files/figure-latex/d_relationship_figure-1.pdf}\\
Figure 2: Plots depicting the function
\(d = cor(x, x_O) - cor(x,y)cor(y, y_O)\).

@@

Figure 3: Outcomes were simulated to be causally influenced by
exposures, with varying degrees of measurement error applied to both.
CIT and MR were used to infer evidence for causality between the
exposure and outcome, and to infer the direction of causality. The value
of \(d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}\), such that when \(d\)
is negative we expect the Steiger test to be more likely to be wrong
about the direction of causality. Rows of graphs represent the sample
size used in the simulations. For the CIT method, outcome 1 denoted
evidence for causality with correct model, outcomes 2 or 3 denoted
evidence for causality with incorrect model, and outcome 4 denoted no
evidence for causality.

@@

\[
\begin{aligned}
x & = \alpha_g + \beta_g g + \epsilon_g \\
x_O & = \alpha_m + \beta_m x + \epsilon_m \\
y & = \alpha_x + \beta_x x + \epsilon_x
\end{aligned}
\]

where \(x\) is the exposure on the outcome \(y\), \(g\) is an instrument
that has a direct effect on \(x\), and \(x_O\) is the measured quantity
of \(x\), where measurement error is incurred from bias in \(\alpha_m\)
and \(\beta_m\) and imprecision from \(\epsilon_m\), our objective is to
estimate the expected magnitude of association between \(g\) and \(y\)
after conditioning on \(x\). Under the CIT, this is expected to be
\(cov(g, y - \hat{y}) = 0\) when \(x\) causes \(y\), where
\(\hat{y} = \hat{a}_{x_O} + \hat{b}_{x_O} x_O + \epsilon_{x_O}\) is the
predicted value of \(y\) using the measured value of \(x_O\).

We can split \(cov(g, y - \hat{y})\) into two parts, \(cov(g, y)\) and
\(cov(g, \hat{y})\).

@@

\[
\begin{aligned}
cov(g, y) & = cov(g, \beta_x x) \\
          & = cov(g, \beta_x \beta_g g) \\
          & = \beta_x\beta_g var(g)
\end{aligned}
\]

@@

\[
\begin{aligned}
cov(g, \hat{y}) & = cov(g, \hat{\beta}_{x_O} x_O) \\
                & = cov(g, \hat{\beta}_{x_O} \beta_m x) \\
                & = cov(g, \hat{\beta}_{x_O} \beta_m \beta_g g) \\
                & = \hat{\beta}_{x_O} \beta_m \beta_g var(g)
\end{aligned}
\]

@@

1.Millstein, J., Zhang, B., Zhu, J. \& Schadt, E. E. Disentangling
molecular relationships with a causal inference test. \emph{BMC
genetics} \textbf{10,} 23 (2009).

>>>>

\setlength{\droptitle}{-2em}
  \title{Determining the existance and direction of causality in the face of
measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{05 May 2016}

@@

* Correspondence to:
\href{mailto:g.hemani@bristol.ac.uk}{g.hemani@bristol.ac.uk}

@@

With the explosion in data that assess the human phenome, we are
presented with the challenge of understanding the causal nature of the
emerging putitive correlations between measures of interest. Mediation
is a commonly used statistical method for making causal inference in
observational data \textsuperscript{1--4}, and it exists in many forms,
from simple regression based systems to structural equation modeling
(SEM) \textsuperscript{5}. The principle behind mediation analysis can
be explained as follows. Supposing an exposure has a causal effect on an
outcome, then an `instrument' that causes the exposure (a single
nucleotide polymorophism (SNP), for example) should also influence the
outcome. Therefore the influence of the instrument on the outcome
conditional on the exposure should be zero. This forms the basis of a
number of methods, such as the regression-based causal inference test
(CIT) \textsuperscript{1}, an SEM implementation in the NEO software
\textsuperscript{5}, and various Bayesian methods, which have been
employed by a number of recent publications that make causal inferences,
often in large scale `omics datasets \textsuperscript{6--8}.

While mediation based approaches are simple and convenient to implement,
it is important to note where they may lead to unreliable results. One
such mechanism is measurement error \textsuperscript{9--12}, which has
been noted to be an explicit issue with CIT style approaches
\textsuperscript{13,14}. Measurement (or observational) error is the
difference between a measured value of a quantity and its true value.
Such variability can arise through a whole plethora of mechanisms, which
are often unique to the study design and difficult to avoid
\textsuperscript{15,16}. Array technology is now commonly used to obtain
high throughput phenotyping at low cost, but they come with the problem
of having imperfect sensitivity, so for example methylation levels as
measured by the Illumina450k chip are prone to have some amount of noise
around the true value \textsuperscript{17,18}. Sensitivity is also an
issue, for example if the unit of measurement of biological interest is
the methylation level in a T cell, then measurement error of this value
can be introduced by using methylation levels from whole blood samples
because the measured value will be an assay of many cell types
\textsuperscript{19}.

Measurement error will of course arise in more low-tech data too, for
example when measuring body mass index (BMI) one is typically interested
in using this as a proxy for obesity, but it is clear that the
correlation between BMI and underlying adiposity is not perfect
\textsuperscript{20}. A similar problem of biological misspecification
is unavoidable in disease diagnosis. Measurement error can also be
introduced after the data has been collected, for example the
transformation of non-normal data for the purpose of statistical
analysis will lead to a new variable that will typically have both bias
and imprecision compared to the original variable. The sources of
measurement error are not limited to this list \textsuperscript{16}, and
its impact has been explored in the context of mediation analysis in the
epidemiological literature extensively \textsuperscript{9,12}.

An alternative statistical method that gained prominence as a solution
to the problem of measurement error is Mendelian randomisation (MR)
\textsuperscript{21,22}. Here an instrumental variable (typically a SNP)
that has a robost, causal association with the exposure is used to stand
for the exposure in an association test with the outcome. If there is no
measurement error in the SNP then the causal association between the
exposure and the outcome can be estimated by scaling the association
between the SNP and the outcome by the association between the SNP and
the exposure. Indeed this approach was in part pioneered (in the form of
instrumental variable (IV) analysis) as a correction for measurement
error \textsuperscript{23}, but it also has the useful property that it
guards against unmeasured confounders driving the association between
exposure and outcome; and if the biological effect of the SNP is
understood then it can also aid against reverse causality driving the
association.

Here, however, arises a potential issue: often the biological effect of
a SNP is not known. If there exists a putative association between two
variables, with a SNP being robustly associated with each, it can be
difficult to determine which of the two variables is subject to the
direct effect of the SNP (i.e., for which of the two variables is the
SNP a valid instrument?). By definition, we expect that if the
association is causal then a SNP for the exposure will be associated
with the outcome, so if the researcher erroneously uses the SNP as an
instrument for the outcome then they are likely to see an apparently
robust causal association of outcome on exposure. Such a situation can
arise in many scenarios. For example genome wide association studies
(GWASs) that identify genetic associations for complex traits are, by
design, hypothesis free and agnostic of genomic function, and it often
takes years of follow up studies to understand the biological nature of
a putative GWAS hit \textsuperscript{24}. To illustrate a trivial
example - the same FTO variant that is famously associated with BMI is
also associated with C-reactive protein (CRP). It is known that this is
because BMI causes CRP \textsuperscript{25}, but supposing we only had
\emph{a priori} knowledge that FTO associated with CRP, and we used it
to instrument CRP in an MR analysis of CRP against BMI, we would infer
that CRP causes BMI. Another situation is where the causal direction
between 'omic measures need to be determined, for example if a DNA
methylation probe is associated with expression of an adjacent gene,
then is a cis-acting SNP an instrument for the DNA methylation level, or
the gene expression level \textsuperscript{8}?

@@

The analysis proceeds by simulating variables that have known causal
effects and valid instruments, and then applying measurement error in
different configurations. We then assess the performance of the CIT test
and MR analysis for a) inference of a causal relationship and b)
inferring the correct direction of causality. Included in the MR
analysis is a simple extension that is used to make inference of the
causal direction when it is unknown which of the correlated variables
the instrument is directly acting upon. All analysis was performed using
the R programming language \textsuperscript{26} and code is made
available at \href{}{github url to go here}.

@@

The CIT method \textsuperscript{1} is implemented in the R package
\emph{R/cit} \textsuperscript{27}. The \emph{cit.cp} function was used
to obtain an omnibus p-value of causality between an exposure $x$ and an
outcome $y$ using a mediator $g$ by making inferences based on a number
of conditions that we examine in the results. To infer the direction of
causality using the CIT method, an omnibus p-value generated by CIT,
$p_{CIT}$, was estimated for the correct model of $x$ causing $y$, and
for incorrect model of $y$ causing $x$. For some significance threshold
$\alpha$, causal direction was inferred based on the following
scenarios:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} > \alpha$ then
  the correct model is accepted
\item
  If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} < \alpha$ then
  the incorrect model is accepted
\item
  If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} > \alpha$ then
  the an alternative causal model is accepted
\item
  If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} < \alpha$ then no
  inference is made
\end{enumerate}

For the purposes of compiling simulation results we use an arbitrary
$\alpha = 0.05$ value.

@@

Supposing there are two correlated variables, $x$ and $y$, with a
variant $g$ associated with both. Assuming no horizontal pleiotropy
(i.e. $g$ only has an effect on one of the variables through the other
variable) it is desirable to know which of the variables it has a direct
influence on. This can be achieved by testing for a difference in the
correlations $\rho_{g, x}$ and $\rho_{g, y}$ using Steiger's Z-test for
correlated correlations within a population \textsuperscript{28}. It is
calculated as

@@

where Fisher's z-transformation is used to obtain
$Z_{g*} = \frac{1}{2} \ln \left ( \frac{1+\rho_{g*}}{1-\rho_{g*}} \right )$,

@@

The $Z$ value is interpreted such that

@@

and a p-value is generated from the $Z$ value to indicate confidence of
difference in correlations $\rho_{gx}$ and $\rho_{gy}$. The test for a
causal association is then obtained by performing two sample MR in the
direction that is predicted to be the correct model based on the Z score
test. For the purposes of compiling simulation results, the test is
deemed to have obtained a robust association if the Z score test has a
p-value $< 0.05$ and the two sample MR analysis has a p-value $< 0.05$,
though in practice we do not advocate using an arbitrary $p$ value
threshold for denoting significance in making causal inference.

Note that the same correlation test approach can be applied to a
two-sample MR \textsuperscript{29} setting, where the Steiger test of
two independent correlations is applied instead of the case detailed
here where correlations share one variable in common. An advantage of
using the Steiger test in the two sample context is that it can compare
correlations in independent samples where sample sizes are different.

@@

Simulations were conducting by creating variables of sample size $n$ for
the exposure $x$, the measured values of the exposure $x_O$, the outcome
$y$, the measured values of the outcome $y_O$ and the instrument $g$. In
all models $x$ causes $y$ and $g$ is an instrument for $x$. Each
variable was simulated such that:

@@

All $\beta$ values were set to 1. Values of $\epsilon_*$ were generated
such that

@@

giving a total of 432 combinations of parameters. Each of these sets of
variables was performed 100 times, and the CIT and MR methods were
applied to each in order to evaluate the causal association of the
simualted variables. Similar patterns of results were obtained for
different values of $cor(g, x)$.

@@

Measurement error of an exposure can be modeled as some transformation
of the true value that leads to the observed value, $x_O=f(x)$. For
example, we can define $f(x) = \alpha_m + \beta_m x + \epsilon_m$, where
$\alpha_m$ and $\beta_m$ influence the bias in the measurement of $x$,
and $\epsilon_m$ represents the imprecision in the measurement of $x$.
Here the true value of the exposure is partially explained by the
genetic instrument, $g$, such that

@@

where $\beta_g$ is the effect of the SNP on the exposure, and
$\epsilon_g$ is the residual value of $x$; and the outcome is partially
explained by the exposure

@@

where $\beta_x$ is the true effect of the exposure on the outcome. In
the causal inference test (CIT), an omnibus p-value is generated from
four hypothesis tests: 1) $g$ is associated with $x$; 2) $g$ is
associated with $y$; 3) $g$ is associated with $x|y$; and 4) $g$ is
independent of $y|x$. The 4th condition is necessary for causal
inference, and can be expressed as $cov(g, y - \hat{y}) = 0$, where
$\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_O$. When measurement error
is introduced we can show using basic covariance properties (Appendix 1)
that

@@

Thus an observational study will find $cov(g, y - \hat{y}) = 0$ when the
true model is causal only when $D = 1$. Therefore, if there is any
measurement error that incurs imprecision (i.e.
$var(\epsilon_m) \neq 0$) then there will remain an association between
$g$ and $y|x$, which is in violation of the the 4th condition of the
CIT. Note that measurement bias alone is insufficient to lead to a
violation of the test statistic assumptions.

We performed simulations to verify that this problem does arise using
the CIT method. Figure 1 shows that when there is no measurement error
in the exposure or outcome variables ($\rho_{x, x_O}=1$) the CIT is
reliable in identifying the correct causal direction. However, as
measurement error increases in the exposure variable, eventually the CIT
is more likely to infer a robust causal association in the wrong
direction. Also of concern here is that increasing sample size does not
solve the issue, indeed it only strengthens the evidence of the
incorrect inference.

@@

When selecting an instrument $g$ that has a direct influence on $x$ and
$x$ is causally related to $y$, one can reason that the influence of $g$
on $y$ is the proportion of the effect of $x$ on $y$ that is explained
by the effect of $g$ on $x$. Hence the effect estimate $\beta_{MR}$ of
$x$ on $y$ is estimated as

@@

thus we obtain an estimate of the effect of $x_O$ on $y$ that is
unbiased by measurement error in $x$.

However, if we do not know whether the SNP $g$ has a direct influence on
$x$ or $y$ then further efforts are required. If $x$ causes $y$ and $g$
is a valid instrument for $x$ then we expect that the Steiger test will
obtain the correct result because the value
$d = \rho_{g, x_O} - \rho_{g, y_O}$ is greater than 0 when the causal
association between $x$ and $y$ is $\rho_{x, y} < 1$. But it can be
shown that in the presence of measurement error,
$d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$ (Appendix 2), thus it is
important to note that under certain conditions when measurement error
is present the Steiger test could make erroneous inference about causal
direction because the value of $d$ is not restricted to being greater
than 0. However, using the MR approach there are two potential
advantages over the CIT. First, we can conduct a formal test for the
direction of causality. Second the correct direction of causality can be
inferred when measurement error in the exposure variable is present, as
long as it is lower than the product of the measurement error in the
outcome and the causal correlation between the exposure and the outcome.
Figure 2 shows that in most cases, especially when the causal effect
between $x$ and $y$ is not very large, the condition of $d > 0$ is
satisfied.

We performed simulations to explore the performance of this approach in
comparison to CIT. Figure 3 shows that, as predicted, when $d < 0$ the
MR analysis is liable to infer the wrong direction of causality, and
that this erroneous result is more likely to occur with increasing
sample size. However, in a large number of cases the MR analysis infers
the wrong direction of causality at a much lower rate than the CIT. When
$d > 0$ is satisfied we observe that in most cases the MR method has
greater power to obtain evidence for causality than CIT, and always
obtains the correct direction of causality.

@@

Under many circumstances a practical solution to this problem is to use
Mendelian randomisation instead of mediation based methods. Mendelian
randomisation is robust in the face of measurement error and, given that
the researcher has knowledge about the biology of the instrument being
used in the analysis, can offer a direct solution to the issues that CIT
faces. This assumption is often reasonable, for example SNPs are
commonly used as instruments when they are found in genes with known
biological relevance to the trait of interest. But on many occasions
this is not the case, and it may be tempting to resort to using methods
based on mediation in order to be able to both ascertain if there is a
causal association and to infer the direction of causality. Here we have
described a simple extension to MR which can be used as an alternative
to mediation based methods. We show that this method is still liable to
measurement error, but because it has different properties to the CIT it
offers two main advantages. First, it uses a formal statistical
framework to test for the robostness of the assumed direction of
causality. Second, after testing in a comprehensive range of scenarios
the MR based approach is less likely to infer the wrong direction of
causality compared to CIT, while substantially improving power over CIT
in the cases where $d > 0$.

Mediation based network approaches are very well established
\textsuperscript{14} and have a number of extensions that make them
valuable tools, including for example network construction. But because
they are predicated on the basic underlying principles of mediation they
are liable to suffer from the same issues of measurement error. Recent
advances in MR methodology, for example network MR \textsuperscript{30}
and mediation through MR \textsuperscript{31} may offer more robust
solutions alternatives for these more complicated problems.

There are a number of limitations to the analysis that we performed. We
focused on the simple case of a single instrument in a single sample
setting, and assumed that pleiotropy (the influence of the instrument on
the outcome through a mechanism other than the exposure) was not
present. To give an example in which this might arise consider the case
that C-reactive protein (CRP) is being tested for causal association
with some outcome, and the FTO variant is used as an instrument. Though
FTO is robustly associated with CRP, it is through the upstream
influence of BMI \textsuperscript{25}. Thus, if BMI has a direct
influence on the outcome, the conditions of MR will be violated and the
Steiger test proposed here will be unreliable.

@@

\includegraphics{manuscript_files/figure-latex/d_relationship_figure-1.pdf}

Figure 2: Plots depicting the function
$d = cor(x, x_O) - cor(x,y)cor(y, y_O)$.

@@

Figure 3: Outcomes were simulated to be causally influenced by
exposures, with varying degrees of measurement error applied to both.
CIT and MR were used to infer evidence for causality between the
exposure and outcome, and to infer the direction of causality. The value
of $d = \rho_{x, x_O} - \rho_{x,y}\rho_{y,y_O}$, such that when $d$ is
negative we expect the Steiger test to be more likely to be wrong about
the direction of causality. Rows of graphs represent the sample size
used in the simulations. For the CIT method, outcome 1 denoted evidence
for causality with correct model, outcomes 2 or 3 denoted evidence for
causality with incorrect model, and outcome 4 denoted no evidence for
causality.

@@

\[
\begin{aligned}
x   & = \alpha_g + \beta_g g + \epsilon_g \\
x_O & = \alpha_m + \beta_m x + \epsilon_m \\
y   & = \alpha_x + \beta_x x + \epsilon_x \\
y_O & = \alpha_n + \beta_n y + \epsilon_n
\end{aligned}
\]

where $x$ is the exposure on the outcome $y$, $g$ is an instrument that
has a direct effect on $x$, $x_O$ is the measured quantity of $x$, where
measurement error is incurred from bias in $\alpha_m$ and $\beta_m$ and
imprecision from $\epsilon_m$, and $y_O$ is the measured quantity of
$y$, where measurement error is incurred from bias in $\alpha_n$ and
$\beta_n$ and imprecision from $\epsilon_n$, our objective is to
estimate the expected magnitude of association between $g$ and $y$ after
conditioning on $x$. Under the CIT, this is expected to be
$cov(g, y_O - \hat{y_O}) = 0$ when $x$ causes $y$, where
$\hat{y}_O = a_n + \beta_n(\hat{a}_{x_O} + \hat{\beta}_{x_O} x_O)$ is
the predicted value of $y_O$ using the measured value of $x_O$.

We can split $cov(g, y_O - \hat{y}_O)$ into two parts, $cov(g, y_O)$ and
$cov(g, \hat{y}_O)$.

@@

\[
\begin{aligned}
cov(g, y_O) & = cov(g, \beta_n y) \\
            & = cov(g, \beta_n \beta_x x) \\
            & = cov(g, \beta_n \beta_x \beta_g g) \\
            & = \beta_n \beta_x \beta_g var(g)
\end{aligned}
\]

@@

\[
\begin{aligned}
cov(g, \hat{y}_O) & = cov(g, \alpha_n + \beta_n \alpha_x + \beta_n \hat{\beta}_{x_O} x_O) \\
                  & = cov(g, \beta_n \hat{\beta}_{x_O} \beta_m x) \\
                  & = cov(g, \beta_n \hat{\beta}_{x_O} \beta_m \beta_g g) \\
                  & = \hat{\beta}_{x_O} \beta_n \beta_m \beta_g var(g)
\end{aligned}
\]

@@

1.Millstein, J., Zhang, B., Zhu, J. \& Schadt, E. E.\emph{et al.}
Disentangling molecular relationships with a causal inference test.
\emph{BMC genetics} \textbf{10,} 23 (2009).
